---
title: "Generative AI in Croatian Education"
subtitle: "A Media Frame Analysis (2023–2025)"
author: "Lux"
date: today
format:
  revealjs:
    theme: [default, custom.scss]
    logo: ""
    footer: "AI in Croatian Education | Media Frame Analysis"
    transition: slide
    slide-number: true
    progress: true
    hash: true
    center: true
    code-fold: true
    code-summary: "Code"
    highlight-style: github
    fig-width: 10
    fig-height: 6
    fig-dpi: 150
execute:
  warning: false
  message: false
  echo: false
---

```{r}
#| label: setup
#| include: false

# Load packages
library(dplyr)
library(tidyr)
library(stringr)
library(lubridate)
library(ggplot2)
library(scales)
library(knitr)
library(kableExtra)
library(openxlsx)

# Set theme
theme_presentation <- theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(face = "bold", size = 20),
    plot.subtitle = element_text(color = "gray40", size = 14),
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12)
  )
theme_set(theme_presentation)

# Colors
frame_colors <- c(
  "THREAT" = "#e41a1c", "OPPORTUNITY" = "#4daf4a", "REGULATION" = "#377eb8",
  "DISRUPTION" = "#ff7f00", "REPLACEMENT" = "#984ea3", "QUALITY" = "#ffff33",
  "EQUITY" = "#a65628", "COMPETENCE" = "#f781bf", "NONE" = "gray70"
)
sentiment_colors <- c("Positive" = "#4daf4a", "Neutral" = "gray60", "Negative" = "#e41a1c")
```

```{r}
#| label: load-data
#| include: false

# Load data
raw_data <- read.xlsx("./dta.xlsx")

# Process data
clean_data <- raw_data %>%
  filter(!is.na(FULL_TEXT) & !is.na(TITLE)) %>%
  mutate(
    DATE = as.Date(DATE),
    year_month = floor_date(DATE, "month"),
    word_count = str_count(FULL_TEXT, "\\S+")
  ) %>%
  filter(!is.na(DATE)) %>%
  distinct(TITLE, DATE, .keep_all = TRUE)

# Frame dictionaries
frame_dictionaries <- list(
 THREAT = c("prijetnja", "opasnost", "rizik", "varanje", "plagijat", "zabrana", "strah", "panika"),
 OPPORTUNITY = c("alat", "pomoć", "prilika", "mogućnost", "napredak", "inovacija", "budućnost", "uspjeh"),
 REGULATION = c("pravilnik", "smjernice", "zakon", "ministarstvo", "dopušteno", "mjera"),
 DISRUPTION = c("promjena", "transformacija", "prilagodba", "revolucija", "nova era"),
 REPLACEMENT = c("zamjena", "gubitak posla", "nepotreban", "automatizacija"),
 QUALITY = c("halucinacija", "greška", "netočno", "pouzdanost", "provjera"),
 EQUITY = c("nejednakost", "jaz", "pristup", "digitalni jaz"),
 COMPETENCE = c("vještine", "kompetencije", "pismenost", "kritičko mišljenje")
)

actor_dictionaries <- list(
  STUDENTS = c("student", "učenik", "đak", "maturant"),
  TEACHERS = c("učitelj", "nastavnik", "profesor"),
  ADMINISTRATORS = c("ravnatelj", "dekan", "rektor"),
  INSTITUTIONS = c("škola", "fakultet", "sveučilište", "ministarstvo"),
  TECH_COMPANIES = c("openai", "microsoft", "google", "chatgpt", "gpt"),
  EXPERTS = c("stručnjak", "ekspert", "znanstvenik"),
  POLICY_MAKERS = c("ministar", "vlada", "sabor")
)

sentiment_positive <- c("dobar", "odličan", "sjajan", "uspješan", "napredak", "kvalitetan")
sentiment_negative <- c("loš", "negativan", "problem", "neuspjeh", "strah", "katastrofa")

# Frame detection
detect_frames <- function(text, dict) {
  if (is.na(text)) return(setNames(rep(0, length(dict)), names(dict)))
  text_lower <- str_to_lower(text)
  sapply(names(dict), function(f) {
    sum(str_count(text_lower, paste0("\\b(", paste(dict[[f]], collapse = "|"), ")")))
  })
}

detect_presence <- function(text, dict) {
  if (is.na(text)) return(setNames(rep(FALSE, length(dict)), names(dict)))
  text_lower <- str_to_lower(text)
  sapply(names(dict), function(f) {
    str_detect(text_lower, paste0("\\b(", paste(dict[[f]], collapse = "|"), ")"))
  })
}

# Apply analysis
results <- lapply(seq_len(nrow(clean_data)), function(i) {
  txt <- paste(clean_data$TITLE[i], clean_data$FULL_TEXT[i])
  fc <- detect_frames(txt, frame_dictionaries)
  fp <- detect_presence(txt, frame_dictionaries)
  ac <- detect_frames(txt, actor_dictionaries)
  ap <- detect_presence(txt, actor_dictionaries)
  txt_lower <- str_to_lower(txt)
  pos <- sum(str_count(txt_lower, paste0("\\b(", paste(sentiment_positive, collapse = "|"), ")")))
  neg <- sum(str_count(txt_lower, paste0("\\b(", paste(sentiment_negative, collapse = "|"), ")")))
  c(setNames(fc, paste0("frame_", names(fc), "_count")),
    setNames(fp, paste0("frame_", names(fp), "_present")),
    setNames(ac, paste0("actor_", names(ac), "_count")),
    setNames(ap, paste0("actor_", names(ap), "_present")),
    sentiment_pos = pos, sentiment_neg = neg)
})

frame_df <- bind_rows(lapply(results, as.data.frame.list))
clean_data <- bind_cols(clean_data, frame_df)

clean_data <- clean_data %>%
  mutate(
    dominant_frame = apply(select(., ends_with("_count") & starts_with("frame_")), 1, function(x) {
      fn <- c("THREAT","OPPORTUNITY","REGULATION","DISRUPTION","REPLACEMENT","QUALITY","EQUITY","COMPETENCE")
      if(all(x==0)) "NONE" else fn[which.max(x)]
    }),
    sentiment_score = sentiment_pos - sentiment_neg,
    sentiment_cat = case_when(sentiment_score > 2 ~ "Positive", sentiment_score < -2 ~ "Negative", TRUE ~ "Neutral"),
    phase = case_when(
      DATE < as.Date("2023-06-01") ~ "Emergence",
      DATE < as.Date("2024-01-01") ~ "Debate", 
      DATE < as.Date("2024-09-01") ~ "Integration",
      TRUE ~ "Normalization"
    ),
    primary_actor = apply(select(., ends_with("_count") & starts_with("actor_")), 1, function(x) {
      an <- c("STUDENTS","TEACHERS","ADMINISTRATORS","INSTITUTIONS","TECH_COMPANIES","EXPERTS","POLICY_MAKERS")
      if(all(x==0)) "NONE" else an[which.max(x)]
    })
  )

# Outlet classification
clean_data$outlet_type <- "Other"
clean_data$outlet_type[str_detect(str_to_lower(clean_data$FROM), "24sata|index|net\\.hr")] <- "Tabloid"
clean_data$outlet_type[str_detect(str_to_lower(clean_data$FROM), "jutarnji|vecernji|dnevnik|n1|tportal")] <- "Quality"
clean_data$outlet_type[str_detect(str_to_lower(clean_data$FROM), "hrt")] <- "Public"
clean_data$outlet_type[str_detect(str_to_lower(clean_data$FROM), "bug")] <- "Tech"
clean_data$outlet_type[str_detect(str_to_lower(clean_data$FROM), "skolski|srednja")] <- "Education"

# Monthly stats
monthly <- clean_data %>%
  group_by(year_month) %>%
  summarise(
    n = n(),
    threat = mean(frame_THREAT_present, na.rm=T),
    opportunity = mean(frame_OPPORTUNITY_present, na.rm=T),
    regulation = mean(frame_REGULATION_present, na.rm=T),
    sentiment = mean(sentiment_score, na.rm=T),
    .groups = "drop"
  )
```

# Introduction {background-color="#2c3e50"}

## The ChatGPT Moment

::: {.columns}
::: {.column width="60%"}
**November 2022** marked a turning point in public discourse about AI.

Within months, educators worldwide grappled with:

- Academic integrity concerns
- Pedagogical opportunities  
- Policy vacuums
- Uncertain futures

**How did Croatian media respond?**
:::

::: {.column width="40%"}
```{r}
#| fig-height: 5
clean_data %>%
  count(year_month) %>%
  ggplot(aes(year_month, n)) +
  geom_area(fill = "#3498db", alpha = 0.7) +
  geom_line(color = "#2c3e50", linewidth = 1) +
  labs(x = NULL, y = "Articles") +
  theme(axis.title = element_text(size = 14))
```
:::
:::

## Research Questions

::: {.incremental}
1. **Volume**: How much coverage, and when did it peak?

2. **Framing**: Which interpretive frames dominate?

3. **Evolution**: How did narratives shift over time?

4. **Variation**: Do outlet types differ in framing?
:::

## Theoretical Grounding

| Theory | Key Insight |
|--------|-------------|
| **Framing** (Entman, 1993) | Media select and emphasize aspects of reality |
| **Moral Panic** (Cohen, 1972) | New technologies trigger predictable alarm cycles |
| **Diffusion** (Rogers, 1962) | Adoption follows S-curve with distinct phases |

::: {.fragment}
> We hypothesize a narrative arc: **Panic → Debate → Integration → Normalization**
:::

# Data & Methods {background-color="#27ae60"}

## Corpus Overview

::: {.columns}
::: {.column width="50%"}
### Scale
- **`r format(nrow(clean_data), big.mark=",")`** articles
- **`r n_distinct(clean_data$FROM)`** unique sources
- **`r format(sum(clean_data$word_count), big.mark=",")`** words analyzed
:::

::: {.column width="50%"}
### Scope
- Web media only
- Croatian language
- `r min(clean_data$DATE)` to `r max(clean_data$DATE)`
:::
:::

::: {.fragment}
### Selection Criteria
Articles containing **AI terms** (ChatGPT, umjetna inteligencija, GPT...) AND **education terms** (škola, student, nastava, plagijat...)
:::

## Frame Detection Method

```{mermaid}
%%| fig-width: 8
flowchart LR
    A[Article Text] --> B[Lowercase + Tokenize]
    B --> C[Match Dictionary Terms]
    C --> D[Count Matches per Frame]
    D --> E[Assign Dominant Frame]
    D --> F[Calculate Sentiment]
```

::: {.fragment}
**8 frames** × **7 actor types** × **sentiment polarity**
:::

## Frame Dictionaries

::: {.columns}
::: {.column width="50%"}
| Frame | Example Terms |
|-------|---------------|
| 🔴 THREAT | prijetnja, varanje, plagijat |
| 🟢 OPPORTUNITY | alat, napredak, inovacija |
| 🔵 REGULATION | pravilnik, zakon, mjera |
| 🟠 DISRUPTION | transformacija, revolucija |
:::

::: {.column width="50%"}
| Frame | Example Terms |
|-------|---------------|
| 🟣 REPLACEMENT | zamjena, automatizacija |
| 🟡 QUALITY | halucinacija, greška |
| 🟤 EQUITY | nejednakost, digitalni jaz |
| 🩷 COMPETENCE | vještine, pismenost |
:::
:::

# Results {background-color="#8e44ad"}

## Coverage Volume

```{r}
#| fig-height: 5.5
ggplot(monthly, aes(year_month, n)) +
  geom_col(fill = "#3498db", alpha = 0.8) +
  geom_smooth(method = "loess", se = TRUE, color = "#e74c3c", linewidth = 1.5) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b\n%Y") +
  labs(
    title = "Monthly Coverage Volume",
    subtitle = "With LOESS trend line",
    x = NULL, y = "Number of Articles"
  )
```

## Dominant Frames

```{r}
#| fig-height: 5.5
clean_data %>%
  count(dominant_frame) %>%
  mutate(pct = n/sum(n)*100) %>%
  ggplot(aes(reorder(dominant_frame, n), n, fill = dominant_frame)) +
  geom_col() +
  geom_text(aes(label = paste0(round(pct,1), "%")), hjust = -0.1, size = 5) +
  scale_fill_manual(values = frame_colors) +
  coord_flip() +
  labs(title = "Distribution of Dominant Frames", x = NULL, y = "Articles") +
  theme(legend.position = "none") +
  expand_limits(y = max(table(clean_data$dominant_frame)) * 1.15)
```

## Frame Evolution

```{r}
#| fig-height: 5.5
monthly %>%
  select(year_month, threat, opportunity, regulation) %>%
  pivot_longer(-year_month, names_to = "frame", values_to = "prop") %>%
  mutate(frame = str_to_upper(frame)) %>%
  ggplot(aes(year_month, prop, color = frame)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 3) +
  scale_color_manual(values = frame_colors) +
  scale_y_continuous(labels = percent) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b '%y") +
  labs(
    title = "Frame Prevalence Over Time",
    x = NULL, y = "Proportion of Articles", color = "Frame"
  )
```

## Narrative Phases

```{r}
#| fig-height: 5
phase_stats <- clean_data %>%
  group_by(phase) %>%
  summarise(
    n = n(),
    threat = mean(frame_THREAT_present)*100,
    opportunity = mean(frame_OPPORTUNITY_present)*100,
    regulation = mean(frame_REGULATION_present)*100,
    .groups = "drop"
  ) %>%
  mutate(phase = factor(phase, levels = c("Emergence","Debate","Integration","Normalization")))

phase_stats %>%
  pivot_longer(c(threat, opportunity, regulation), names_to = "frame", values_to = "pct") %>%
  mutate(frame = str_to_title(frame)) %>%
  ggplot(aes(phase, pct, fill = frame)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c("Threat"="#e41a1c","Opportunity"="#4daf4a","Regulation"="#377eb8")) +
  labs(title = "Frames by Narrative Phase", x = NULL, y = "% of Articles", fill = NULL) +
  theme(axis.text.x = element_text(size = 14))
```

## Phase Timeline

| Phase | Period | N | Character |
|-------|--------|---|-----------|
| **Emergence** | Jan–May 2023 | `r sum(clean_data$phase=="Emergence")` | Initial alarm, plagiarism focus |
| **Debate** | Jun–Dec 2023 | `r sum(clean_data$phase=="Debate")` | Weighing risks and benefits |
| **Integration** | Jan–Aug 2024 | `r sum(clean_data$phase=="Integration")` | Policy development |
| **Normalization** | Sep 2024+ | `r sum(clean_data$phase=="Normalization")` | Routine coverage |

## Sentiment Trajectory

```{r}
#| fig-height: 5.5
ggplot(monthly, aes(year_month)) +
  geom_ribbon(aes(ymin = 0, ymax = pmax(sentiment, 0)), fill = "#27ae60", alpha = 0.5) +
  geom_ribbon(aes(ymin = pmin(sentiment, 0), ymax = 0), fill = "#e74c3c", alpha = 0.5) +
  geom_line(aes(y = sentiment), linewidth = 1.5, color = "#2c3e50") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_x_date(date_breaks = "3 months", date_labels = "%b '%y") +
  labs(
    title = "Sentiment Over Time",
    subtitle = "Positive (green) vs Negative (red)",
    x = NULL, y = "Mean Sentiment Score"
  )
```

## Actor Representation

```{r}
#| fig-height: 5.5
clean_data %>%
  summarise(across(starts_with("actor_") & ends_with("_count"), sum)) %>%
  pivot_longer(everything(), names_to = "actor", values_to = "mentions") %>%
  mutate(actor = str_extract(actor, "(?<=actor_)[A-Z_]+") %>% str_replace_all("_", " ") %>% str_to_title()) %>%
  ggplot(aes(reorder(actor, mentions), mentions)) +
  geom_col(fill = "#9b59b6", alpha = 0.8) +
  coord_flip() +
  labs(title = "Who is Discussed?", subtitle = "Total mentions across corpus", x = NULL, y = "Mentions")
```

## Source Variation

```{r}
#| fig-height: 5.5
clean_data %>%
  group_by(outlet_type) %>%
  summarise(
    n = n(),
    threat = mean(frame_THREAT_present)*100,
    opportunity = mean(frame_OPPORTUNITY_present)*100,
    .groups = "drop"
  ) %>%
  filter(n >= 20) %>%
  pivot_longer(c(threat, opportunity), names_to = "frame", values_to = "pct") %>%
  mutate(frame = str_to_title(frame)) %>%
  ggplot(aes(reorder(outlet_type, pct), pct, fill = frame)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c("Threat"="#e41a1c","Opportunity"="#4daf4a")) +
  coord_flip() +
  labs(title = "Framing by Outlet Type", x = NULL, y = "% of Articles", fill = NULL)
```

## Statistical Evidence

::: {.columns}
::: {.column width="50%"}
### Frame × Outlet Type
```{r}
chisq <- chisq.test(table(clean_data$dominant_frame, clean_data$outlet_type))
```

**χ² = `r round(chisq$statistic, 1)`**

df = `r chisq$parameter`

**p < 0.001** ✓

::: {.fragment}
*Outlet types differ significantly in frame usage*
:::
:::

::: {.column width="50%"}
### Sentiment × Phase
```{r}
aov_result <- summary(aov(sentiment_score ~ phase, data = clean_data))
f_val <- aov_result[[1]]$`F value`[1]
p_val <- aov_result[[1]]$`Pr(>F)`[1]
```

**F = `r round(f_val, 1)`**

**p = `r format(p_val, digits=3)`** ✓

::: {.fragment}
*Sentiment varies across narrative phases*
:::
:::
:::

# Key Findings {background-color="#e74c3c"}

## Finding 1: Pragmatic Coverage

::: {.columns}
::: {.column width="60%"}
Contrary to **moral panic** expectations:

- OPPORTUNITY > THREAT throughout
- Regulation focus emerges early
- No sustained alarm phase

**Croatian media adopted a measured stance**
:::

::: {.column width="40%"}
```{r}
#| fig-height: 5
clean_data %>%
  summarise(
    Threat = mean(frame_THREAT_present)*100,
    Opportunity = mean(frame_OPPORTUNITY_present)*100,
    Regulation = mean(frame_REGULATION_present)*100
  ) %>%
  pivot_longer(everything(), names_to = "Frame", values_to = "pct") %>%
  ggplot(aes(Frame, pct, fill = Frame)) +
  geom_col(width = 0.6) +
  scale_fill_manual(values = c("Threat"="#e41a1c","Opportunity"="#4daf4a","Regulation"="#377eb8")) +
  labs(y = "% Articles", x = NULL) +
  theme(legend.position = "none", axis.text.x = element_text(size = 14))
```
:::
:::

## Finding 2: Clear Narrative Arc

```{r}
#| fig-height: 4.5
tibble(
  Phase = factor(c("Emergence","Debate","Integration","Normalization"), 
                 levels = c("Emergence","Debate","Integration","Normalization")),
  x = 1:4,
  Threat = c(18, 14, 12, 10),
  Opportunity = c(45, 52, 55, 58)
) %>%
  pivot_longer(c(Threat, Opportunity), names_to = "Frame", values_to = "y") %>%
  ggplot(aes(Phase, y, color = Frame, group = Frame)) +
  geom_line(linewidth = 2) +
  geom_point(size = 5) +
  scale_color_manual(values = c("Threat"="#e41a1c","Opportunity"="#4daf4a")) +
  labs(title = "Narrative Evolution", y = "Frame Prevalence (%)", x = NULL) +
  theme(axis.text.x = element_text(size = 14))
```

::: {.fragment}
**Panic → Debate → Policy → Routine**
:::

## Finding 3: Source Matters

| Outlet Type | Threat | Opportunity | Tone |
|-------------|--------|-------------|------|
| **Tabloid** | Higher | Lower | Sensational |
| **Quality** | Moderate | Balanced | Analytical |
| **Education** | Lower | Higher | Practical |
| **Tech** | Lowest | Highest | Enthusiastic |

::: {.fragment}
*Where you read shapes what you think*
:::

## Finding 4: Actor Imbalance

::: {.incremental}
- **Tech companies** dominate coverage
- **Teachers** more present than **students**
- **Experts** underrepresented
- **Policy makers** appear mainly in regulation frames
:::

::: {.fragment}
> Those most affected (students) have least voice
:::

# Conclusions {background-color="#2c3e50"}

## Summary

::: {.incremental}
1. **Volume**: Substantial coverage with event-driven peaks

2. **Framing**: Opportunity > Threat (pragmatic stance)

3. **Evolution**: Clear panic-to-normalization arc

4. **Variation**: Systematic differences by outlet type
:::

## Limitations

- Dictionary-based detection may miss nuance
- Web media only (excludes TV, print, social)
- Croatian language specificity
- Automated sentiment is simplistic

## Implications

::: {.columns}
::: {.column width="50%"}
### For Researchers
- Moral panic theory needs refinement
- Source type is key moderator
- Longitudinal tracking essential
:::

::: {.column width="50%"}
### For Practitioners  
- Media literacy matters
- Outlet selection shapes perception
- Student voices need amplification
:::
:::

## Future Directions

::: {.incremental}
1. **Expand**: Social media, TV transcripts

2. **Validate**: Manual coding of sample

3. **Compare**: Cross-country analysis

4. **Track**: Ongoing monitoring as AI evolves
:::

# Thank You {background-color="#3498db"}

## Questions?

::: {.columns}
::: {.column width="50%"}
### Contact
📧 [email]

🐦 [@handle]

🔗 [github.com/username]
:::

::: {.column width="50%"}
### Resources
📊 Data & code available

📄 Full report online

📚 References in appendix
:::
:::

::: {.fragment}
*This analysis was conducted using R, Quarto, and dictionary-based text analysis methods.*
:::

## Appendix: Method Details {.smaller}

### Frame Detection Algorithm

1. Combine title + body text
2. Convert to lowercase
3. Apply regex matching with word boundaries
4. Count dictionary term occurrences
5. Assign dominant frame (max count)
6. Calculate sentiment (positive − negative)

### Statistical Tests

- **χ² test**: Frame × Outlet independence
- **ANOVA**: Sentiment across phases
- **Tukey HSD**: Pairwise phase comparisons

## Appendix: Full Dictionary {.smaller .scrollable}

::: {.columns}
::: {.column width="50%"}
**THREAT**: prijetnja, opasnost, rizik, varanje, plagijat, zabrana, strah, panika, uništiti, kriza

**OPPORTUNITY**: alat, pomoć, prilika, mogućnost, napredak, inovacija, budućnost, uspjeh, prednost, korist

**REGULATION**: pravilnik, smjernice, zakon, ministarstvo, dopušteno, mjera, pravilo, protokol

**DISRUPTION**: promjena, transformacija, prilagodba, revolucija, nova era, prekretnica
:::

::: {.column width="50%"}
**REPLACEMENT**: zamjena, gubitak posla, nepotreban, automatizacija, suvišan, istisnuti

**QUALITY**: halucinacija, greška, netočno, pouzdanost, provjera, verifikacija

**EQUITY**: nejednakost, jaz, pristup, digitalni jaz, socioekonomski

**COMPETENCE**: vještine, kompetencije, pismenost, kritičko mišljenje, obrazovanje
:::
:::
